---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "An awesome title for a great paper"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
crossref: 
  fig-title: Fig.
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      orcid: 0000-0002-4996-0059
      institute: [TBEP]
      correspondence: true
  - Kerry Flaherty-Walia:
      email: kfwalia@tbep.org
      orcid: 0000-0002-6032-0570
      institute: [TBEP]
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      orcid: 0000-0002-3847-6501      
      institute: [TBEP]
      correspondence: false 
  - Edward T Sherwood:
      email: esherwood@tbep.org
      orcid: 0000-0001-5330-302X
      institute: [TBEP]
      correspondence: false

institute:
  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA

filters: 
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  
execute: 
  echo: false
  warning: false
  ft.keepnext: false
  ft.align: left
---

```{r}
#| results: hide
library(knitr)
library(flextable)
library(here)

source(here('R/funcs.R'))

load(file = here('tabs/hydnrmtab.RData'))
```

## Fig

1. observed chl and model fit
1. model fit v normalized
1. grid plots
1. normalized - fit v hydrology

## Table

1. Model summary statistics (include by month?)


## Methods

### Data processing

Salinity gaps were filled using linear interpolation between observed values, where the maximum gap fill was three time steps (months).  Gaps accounted for less than 5% of the observatiosn for each bay segment.  

### GAMs

@fig-obsprd, @fig-prdnrm, @fig-gridplo, @fig-hydnrm

@tbl-hydnrmtab

ACF plots showed minimal autocorrelation of model residuals at lag 1 (all < 0.20 rho).

K values chosen to allow sufficient degress of freedom for each smoother.  Once models were fit, the `gam.check()` function was used to compare the estimated degrees of freedom for each smoother to the maximum degrees of freedom allowed by the k value.  None of the smoothers were within 80% of the maximum degrees of freedom allowed. 

{{< pagebreak >}}

## Figures {.unnumbered}

![Observed and predicted chlorophll-a ($\mu$g/L) for the four major bay segments of Tampa Bay.  Observed and predicted are shown continuously over time (a) and by month (b). Color shading indicates year. Rows are arranged by bay segment with the top labels in the left column showing the bay segment and the right column showing deviance explained.](figs/obsprd.png){#fig-obsprd}

{{< pagebreak >}}

![Annual predictions, normalized estimates, and mean salinity estimates for chlorophyll-a by major bay segment from the separate GAMs. Normalized values (solid lines) are those that remove the salinity influence using the mean result of predictions across a range of salinity values at each time step, whereas the mean salinity values (dashed lines) are annual predictions across a constant mean salinity for each bay segment. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/prdnrm.png){#fig-prdnrm}

{{< pagebreak >}}

![Predicted chlorophyll-a ($\mu$g/L) during the rainy season (June to November) from 2005 to present as related to salinity (ppth) by major bay segment.  Facets are arranged to show how the monthly predictions vary across years and different salnity ranges.  The points show the model predicted results for the observed salinity.  Salinity ranges in each plot are fixed the minimum and maximum observed across years for the month. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/gridplo.png){#fig-gridplo}

{{< pagebreak >}}

![Salinity-normalized chlorophyll-a ($\mu$g/L) related to hydology from 2004 to present in Old Tampa Bay.  GAM-predicted chlorophyll-a (points) and salinity-normalized (lines) are shown for each quarter (a), followed by hydrologic load anomalies (10$^6$ m$^3$ / quarter) as the load minus the long-term average by quarter (b), followed by the relationship between the predicted minus salinity-normalized chlorophyll and the hydrologic load anomalies by quarter (c). Model R$^2$ values are shown for the linear regression fit (lines) in (c).  JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September; OND: October, November, December.](figs/hydnrm.png){#fig-hydnrm}

{{< pagebreak >}}

## Tables {.unnumbered}

```{r}
#| label: tbl-hydnrmtab
#| tbl-cap: !expr 'paste0("Linear model fits between the predicted minus salinity-normalized chlorophyll and the hydrologic load anomalies by quarter, bay segment, and annual periods. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay. JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September; OND: October, November, December. p < 0.05 *; p < 0.005 **.")'
hydnrmtab
```

{{< pagebreak >}}

## Data Availability {.unnumbered}

All data and analysis code for this manuscript is available on GitHub at <https://github.com/tbep-tech/chlhydro-manu>.

## Acknowledgments {.unnumbered}

This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP). We are thankful for constructive comments provided by xxx anonymous reviewers. We are indebted to the community of field and technical staff that have routinely collected and maintained the valuable long-term monitoring data used in this study.

## Statements and Declarations {.unnumbered}

### Author Contributions {.unnumbered}

Conceptualization: Marcus W Beck, Kerry Flaherty-Walia, Maya C Burke, Edward T Sherwood; Data Curation: Marcus W Beck; Methodology: Marcus W Beck, Kerry Flaherty-Walia, Maya C Burke, Edward T Sherwood; Formal analysis and investigation: Marcus W Beck; Writing - original draft preparation: all authors; Writing - review and editing: all authors; Funding acquisition: Edward T Sherwood.

### Ethics Declaration {.unnumbered}

The authors have no competing interests to declare that are relevant to the content of this article.

### Funding {.unnumbered}

This study was funded by EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St. Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.

## References {.unnumbered}