---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "An awesome title for a great paper"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
crossref: 
  fig-title: Fig.
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      orcid: 0000-0002-4996-0059
      institute: [TBEP]
      correspondence: true
  - Kerry Flaherty-Walia:
      email: kfwalia@tbep.org
      orcid: 0000-0002-6032-0570
      institute: [TBEP]
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      orcid: 0000-0002-3847-6501      
      institute: [TBEP]
      correspondence: false 
  - Edward T Sherwood:
      email: esherwood@tbep.org
      orcid: 0000-0001-5330-302X
      institute: [TBEP]
      correspondence: false

institute:
  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA

filters: 
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  
execute: 
  echo: false
  warning: false
  ft.keepnext: false
  ft.align: left
---

```{r}
#| results: hide
library(knitr)
library(flextable)
library(here)

source(here('R/funcs.R'))

load(file = here('tabs/gamtab.RData'))
load(file = here('tabs/gamcrtab.RData'))
load(file = here('tabs/prdnrmtab.RData'))
load(file = here('tabs/liktab.RData'))
```

## Inroduction 

[@testa2025]

Nitrogen loads in Tampa Bay have been reduced substantially since the 1970s due to concerted efforts to quantify and manage external nutrient inputs.  

WRTDS vs GAMs

The objectives of this study were to (1) develop Generalized Additive Models (GAMs) for chlorophyll-a trends for the four major bay segments of Tampa Bay as a function of time, salinity, and total nitrogen loading, (2) quantify the relative influence of salinity changes and nitrogen loading on chlorophyll-a trends, and (3) assess the potential likelihoods of exceeding regulatory criteria under future salinity changes and different loading scenarios. We hypothesize that water quality drivers have changed over time, warranting a shift in thinking about management strategies to focus on dominant factors that influence chlorophyll-a. This shift may be particularly important in upper bay segments where long-term nitrogen loads have changed and residence time may be affected by freshwater inflows and barriers to circulation. Overall, this work provides an in-depth assessment of chlorophyll trends for informing the management of water quality under past and contemporary conditions, with a generalized framework that could be applied in other locations with long-term data.

## Methods

Tampa Bay, located on the west coast of Florida, is a sub-tropical shallow estuary covering 400 m$^2$ (1036 km$^2$) of open water that includes native habitats of seagrasses, macroalgae, oysters, tidal flats, and hard bottom.  Average depth of the bay is 11 ft with a maximum depth of 43 feet in the dredged shipping channels that primarily traverse the longitudinal axis. The bay is divided into four major segments defined by physical and natural boundaries, with Old Tampa Bay (OTB) in the northwest, Hillsborough Bay (HB) in the northeast, Middle Tampa Bay (MTB), and Lower Tampa Bay (LTB) at the southernmost extent that exchanges with the Gulf of Mexico. The watershed is 2200 mi$2$ (5872 km$^2$) and includes an extensive human presence, with 42% of the land as urban or suburban [@SWFWMD25]. Agricultural and mining activities cover about 25% of the watershed, whereas the remainder includes native habitats of hardwood or scrub uplands and freshwater wetlands.  

Residence time varies with inflow conditions, wind, and spatial scale.  In general, baywide estimates range from 75 (Eulerian) to 159 (Lagrangian) days [@burwell00;@morrison11], with substantial variation by location. Spatially, major inflows enter HB via the Hillsborough and Alafia Rivers, MTB via the Little Manatee river, and LTB via the Manatee River. Comprehensive estimates of residence time by bay segment do not exist, although time generally increases from the mouth of the bay towards the more oligohaline segments. LTB that exchanges with the Gulf has the shortest residence time of a few days, whereas OTB and HB have the longest.  However, HB flushes relatively rapidly compared to OTB due to major freshwater inflows and enhanced baroclinic circulation via the main dredge channel that extends from the mouth of the bay to upper HB [@morrison11].  More importantly, OTB has three major bridges with substantial causeways that are known impediments to circulation [@luther2022].  Residence times in OTB have been estimated at roughly three months [@tbep1115].  Throughout, we consider salinity as a rough proxy for residence time, while also acknowledging that it reflects hydrologic inputs, tidal exchange, baroclinic circulation, and other factors that influence hydrodynanmics (e.g., persistent eddies, wind).

Long-term external nutrient loading from multiple sources has decreased overall in Tampa Bay over the last fifty years [@janicki23]. Point sources primarily from wastewater treatment facilities with outfalls to the bay have been substantially reduced following upgrades to advanced tertiery treatment, resulting in an overall reduction of approximately two thirds from the 1970s to present for the entire bay. Conversely, nonpoint sources primarily from stormwater runoff remain the dominant source of nutrients to the bay, particularly in OTB and HB. Loading from nonpoint sources varies substantially between years with changes in rainfall and with more long-term climatic drivers (i.e., El Ni\~no vs La Ni\~na years).  However, a weak but long-term inrease in rainfall over the past fifty years has lead to gradual increases in stormwater loads and, consequently, a freshening of Tampa Bay, particularly in the upper bay segments [@beck24]. Overall, HB receives the largest external nitrogen loads on an annual basis, followed by MTB, OTB, and LTB, respectively.

Reporting and management of external nutrient loads to Tampa Bay have been the responsibility of the Tampa Bay Nitrogen Management Consortium (TBNMC) since the early 2000s.  These efforts are guided by a Reasonable Assurance Plan whereby reporting and compliance assessment is provided to the Florida Department of Environmental Protection.  Regulatory criteria for chlorophyll-a apply to each bay segment, such that consecutive exceedances greater than two years in a row require an assessment of the assimilative capacity for the segment where the exceedance occurred.  The assimilative capacity for each bay segment is currently defined by a maximum annual nitrogen loading beyond which negative water quality outcomes are expected if exceeded.  Chlorophyll and nitrogen loading thresholds for each bay segment were established primarily to maintain a light environment supportive of seagrass growth.  Efforts of the TBNMC to manage external nutrient loads have contributed in part to the long-term recovery of Tampa Bay.  However, current ecology and physical conditions of the bay compared to past conditions from which regulatory thresholds were developed have been the focus of recent discussions by the TBNMC and have also motivated this current work.  Chlorophyll-a and total load thresholds for each bay segment are evaluated for objective three using the GAMs for scenario assessments under expected future changes and management actions.

### Data processing

Long-term monitoring data for Tampa Bay were obtained from the Environmental Protection Commission of Hillsborough County.  This monitoring program has been ongoing since the mid-1970s, with monthly sampling at fixed stations throughout the bay. Surface chlorophyll-a and salinity observations at each station were averaged by month for each bay segment, with salinity being averaged across bottom, mid, and surface depth measurements.  Salinity and chlorophyll gaps were filled using linear interpolation between observed values, where the maximum gap filled was three months.  Salinity gaps accounted for less than 5% of the observations for each bay segment. Chlorophyll gaps were primarily bserved during the beginning of the COVID-19 pandemic in 2020 when sampling did not occur and only accounted for 2 to 3 observations per bay segment.  Monthly data from 1985 to present by bay segment were used for analysis to coincide with the period of record for the nutrient loading data. 

Monthly total nitrogen loading data to each bay segment from 1985 to present were estimated following methods in @tbep0494 [additional details in @tbep0623].  In brief, loading data were estimated as tons per month based on external source, including domestic point-sources (wastewater treatment effluent, reclaimed water), industrial point sources, non-point sources, atmospheric deposition, groundwater/springs, and material losses from nutrients entering the bay through port/shipping activities.  Methods for calculating point source loads are straightforward and are derived from reported flow and concentration data for every permitted entity in the Tampa Bay watershed. Conversely, non-point source loads are estimated from gauged and ungauged portions of the watershed by combining rainfall and flow estimates with event mean concentration data by land use and soil type.  Atmospheric deposition is estimated using rainfall and airshed chemical constituent data obtained from the Verna well-field long-term air quality monitoring site.  Groundwater loading is estimated from aquifer concentration data combined with potentiometric sub-surface flow data, wheras springs loadings are calculated dircectly from discharge and concentration data measured at each location.  Finally, material losses are based on a loss factor per unit shipped at each port location.  For the purpose of this analysis, no distinction is made between load sources such that the total load as the sum across sources was used for modelling.  However, it is noted that trends by sources have varied by location, e.g., non-point source loads have increased in OTB, whereas point sources have decreased [@janicki23].

### GAMs

Four separate GAMs were created for each bay segment using the combined water quality and loading data for the monthly observations from 1985 to present.  The primary goal was to develop the best fitting model for observed chlorophyll-a by including the effects of salinity, nitrogen loading, and time.  Time was expressed as two separate variables for each time step, first as date expressed as decimal time (e.g., July 1st 2017 is 2017.5) and as day of year to capture a seasonal component.  Each predictor was evaluated as an additive smooth to quantify an individual assocation with chlorophyll-a, whereas interaction smoothers where also included for every pairwise combination of the predictors. No efforts were made for variable selection since the same model structure for describing drivers and scenario testing between bay segments was desired (objectives 2 and 3).  This also allowed direct comparability between models regarding explained variance, either as a whole or for individual predictors.

The model notation was:

$$
\begin{align}
\text{chla}_i &\sim \text{Gamma}(\mu_i, \phi) \\
\log(\mu_i) &= \beta_0 + f_1(\text{dec\_time}_i) + f_2(\text{doy}_i) + f_3(\text{sal}_i) + f_4(\text{load}_i) + \\
&\quad + f_{12}(\text{dec\_time}_i, \text{doy}_i) \\
&\quad + f_{13}(\text{dec\_time}_i, \text{sal}_i) \\
&\quad + f_{14}(\text{dec\_time}_i, \text{load}_i) \\
&\quad + f_{23}(\text{doy}_i, \text{sal}_i) \\
&\quad + f_{24}(\text{doy}_i, \text{load}_i) \\
&\quad + f_{34}(\text{sal}_i, \text{load}_i)
\end{align}
$$

where chlorophyll-a is modelled as a Gamma-distributed response variable with mean $\mu_i$ and dispersion parameter ϕ for observation i (as month).  The decision to use a Gamma distribution as compared to a Gaussian with a log-link function was made given that the variance likely increases with the mean, which is a reasonable expectation for chlorophyll data. The log-transformed mean chlorophyll-a is modelled as an intercept ($\beta_0$) plus smooth functions of decimal time (dec_time), day of year (doy), salinity (sal), and loading (load), along with tensor product smooth interactions between each pair of the four predictors.  Specifically, $f_1​()$ is a thin plate regression spline with k = 40 basis functions, $f_2()$ is a cyclic cubic regression spline with k = 10 basis functions, $f_3()$ and $f_4()$ are thin plate regression splines with k = 10 basis functions, and $f_{12}()$, $f_{13}()$, $f_{14}()$, $f_{23}()$, $f_{24}()$, and $f_{34}()$ are tensor product smooths (excluding main effects) with k = (5, 5) basis functions.  Smoothing parameters were estimated using restricted maximum likelihood [@wood11] using the `gam()` function from the mgcv R package [@wood17].

The K values (number of basis functions) for each smoother were chosen to allow sufficient degress of freedom to describe the assocation of each predictor with chlorophyll.  Once models were fit, the `gam.check()` function was used to compare the estimated degrees of freedom for each smoother to the maximum degrees of freedom allowed by the K value.  None of the smoothers were within 80% of the maximum degrees of freedom allowed, suggesting sufficient flexibility to describe the assocation. ACF plots [@venables2002b] also showed minimal autocorrelation of model residuals at lag 1 (all < 0.20 rho), indicating a time-correlated variance structure was not needed for the models.

In addition to predicted chlorophyll-a, the GAMs can be used to created two additional types of predictions.  First, salinity-normalized predictions were made by predicting chlorophyll-a across a range of salinity values (minimum to maximum observed) at each time step, then taking the mean of those predictions as the salinity-normalized value for that time step.  This produces a time series that can be considered the chlorophyll trend independent of changes in salinity (i.e., just time and loading).  The approach is consistent with other studies examining the effects of salinity on chlorophyll-a concentrations [@beck17;@murphy19].

$$
\begin{align}
\text{Norm}(t) &= \frac{1}{|\mathcal{S}|} \sum_{s \in \mathcal{S}} \hat{f}\left(t, s, \bar{\text{sal}}_{m(t),y(t)}\right)
\end{align}
$$

The final set of predictions for each bay segment were salinity-normalized predictions at mean loading, which can be considered the chlorophyll trend independent of changes in both salinity and loading.  This is accomplished by predicting chlorophyll-a at each time step using the mean loading with the salinity-normalized predictions for each bay segment across the entire time series.  Mean loading is used in this case as opposed to the more involved grid based appraoch for salinity normalization.  This was preferred for ease of interpretation and absence of a typical seasonal cycle that would be expected for a variable like salinity.

Finally, the models for each bay segment were evaluated for fit using conventional summary statistics for GAMs. These statistics included Generalized Cross-Validation scores and explained deviance, as well as estimated and reference degrees of freedom, F, and p-values for each smoother.  Simple correlations between observed and predicted chlorophyll were also evaluated for different year groupings (1985-1999, 2000-2014, 2015-2024) and seasonal periods (annual, JFM, AMJ, JAS, OND) to assess when model fit may differ relative to the entire time series.

### Partitioning Drivers of Chlorophyll Change

The second objective was to assess the relative effects of salinity and loading on chlorophyll trends from the GAMs for each bay segment. This was accomplished by comparing the predicted values from the GAMs with those from the salinity-normalized predictions and the salinity-normalized at mean loading predictions. Specifically, the difference between the predictions and salinity-normalized values as a simple subtraction was used an expression of salinity effects on chlorophyll.  Likewise, the difference between the salinity-normalized values and the salinity-normalized values at mean loading was used an expression of the loading effects on chlorophyll.  

The differences among the predictions by bay segment were assessed as simple visualizations of the annual results over time and using a state-space diagram plotting the differences for each time series over time on separate axes.  For the latter, the x-axis showed the salinity effects (positive being more oligohalien, negative being more euhaline) and the y-axis showed the loading effects (positive being high loading associated with higher chlorophyll, negative being lower loading associated with lower chlorophyll).  Individual years were plotted on the diagram to assess the relative effects of salinity changes and/or loading on the relative chlorophyll changes each year.  Depending on where a year is located on the plot, conclusions can be made on which factor (or both) may or may not have influences chlorophyll for that time period.  Additionally, these differences were further evaluated for different year groupings (1985-1999, 2000-2014, 2015-2024) and seasonal periods (annual, JFM, AMJ, JAS, OND) to further evaluate periods of time when salinity and/or loading had differing effects on chlorophyll.

### Likelihood of Exceeding Regulatory Criteria

The third objective was to use the GAMs to assess the likelihood of exceeding regulatory criteria under projected (future) salinity changes and loading scenarios.  Generally, this was accomplished by repeated simulations from each GAM for the different scenarios.  The `simulate()` function from the gratia R package [@simpson25] was used, which predicts randomized GAM outputs within the prediction interval of the model.  For each scenario, 10,000 simulations were used to estimate chlorophyll, where the mean (and standard deviation) of the percent of years exceeding the thresholds were used to identify the likelihood of exceeding regulatory criteria over time.  Herein, "over time" is defined as a hypothetical period from present to fifty years in the future at one year time steps based on potential salinity changes and loading values for each bay segment. Chlorophyll thresholds for each bay segment are 9.3, 15.0, 8.5, and 5.1 $\mu$g/L and external loading thresholds are 486, 1451, 799, 349 tons per year for OTB, HB, MTB, and LTB, respectively. These thresholds were defined as being supportive of seagrass by promoting a sufficient light environment for growth [@tbep0696;@tbep0400;@tbep0310].

Salinity changes were estimated using the long-term slope of a linear regression from 1975 to present for the annual average salinities for each bay segment, i.e., to obtain estimated change in ppth per year.  Loading scenarios were assessed under four different conditions - current load, loading at the TMDL threshold, half the threshold, and twice the threshold.  The simulations and scenarios were applied to specific blocks of time in the observed period of record for each bay segment, e.g., 2010 to present.Changes in salinity were implemented for the chosen time period at one year time steps up to "50 years" using the expected change per year, where the observed salinity time series was multipled by the slope and the numbers of years in the "future" (one to 50).  Similarly, loading scenarios were evaluated for the four different conditions described above using the observed loading or scaled accordingly within each year, such that the sum of monthly loadings each year was equal to the annual sum for each relative scenario (i.e., half the TDML, equal to TMDL, or double the TMDL).  Practically, the results can be interpreted as the relative effects of potential salinity and loading changes if conditions within the chosen year period were to persist for fifty years into the future.

## Results

@fig-obsprd, @fig-prdnrm, @fig-gridplo, @fig-salvload, @fig-simplo, @fig-simex

@tbl-gamtab, @tbl-gamcrtab, @tbl-prdnrmtab, @tbl-liktab

## Discussion

Why was this particular GAM structure chosen, e.g., decimal time and doy as opposed to decimal time with large k?

Why do chlorophyll simulations using actual loads at time zero show lower likelihoods of exceeding the threshold than the actual percent of years in the time period that exceeded?  For example, in OTB from 2017-2021, 4 of the 5 years exceeded the threshold, producing an expectation that the baseline likelihood would be 80% for the actual load at time zero (observed salinity).  First, the actual values for chlorophyll were very close to the threshold.  Given the inherent uncertainty in measured data (observational and process uncertainy) and the uncertainty of the models (model uncertainty, affected by both observation and process uncertainty), a lower than expected likelihood of exceeding the threshold for that time period was provided by the simulations.  Further, annual exceedances are largely driven by high values during the summer months which produce a higher mean value in a variable that is inherently log-distributed.  Because GAMs, like most empirical models, describe the mean response, these high values are poorly described and are not common outputs in the simulated values.  So, the likelihoods of exceeding the threshold can be considered underestimates, whereas the more important result is the relative increase in the likelihood over time given projected salinity changes in the future.  In other words, all scenarios suggested greater exceedances of regulatory thresholds from present to fifty years.

{{< pagebreak >}}

## Figures {.unnumbered}

![Observed and predicted chlorophll-a ($\mu$g/L) for the four major bay segments of Tampa Bay.  Observed and predicted are shown continuously over time (a) and by month (b). Color shading indicates year. Rows are arranged by bay segment with the top labels in the left column showing the bay segment and the right column showing deviance explained.](figs/obsprd.png){#fig-obsprd}

{{< pagebreak >}}

![Annual predictions, normalized estimates, and normalized estimates at mean loading for chlorophyll-a by major bay segment from the separate GAMs. Normalized values (solid lines) are those that remove the salinity influence using the mean result of predictions across a range of salinity values at each time step, whereas the normalized, mean load values (dashed lines) are salinity-normalized values at a constant mean load for each bay segment. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/prdnrm.png){#fig-prdnrm}

{{< pagebreak >}}

![State space representations of modelled annual chlorophyll response per unit change in salinity (x-axis) or nitrogen load (y-axis).  Points represent values for a given year in (a), joined sequentially through time by the grey lines. Means and 95% confidence for the year groups are shown in (b). OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/salvload.png){#fig-salvload}

{{< pagebreak >}}

![Predicted chlorophyll-a ($\mu$g/L) during the rainy season (June to November) from 2005 to present as related to salinity (ppth) by major bay segment.  Facets are arranged to show how the monthly predictions vary across years and different salnity ranges.  The points show the model predicted results for the observed salinity.  Salinity ranges in each plot are fixed the minimum and maximum observed across years for the month. Grey lines show the chlorophyll contour value for the regulatory threshold for each bay segment. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/gridplo.png){#fig-gridplo}

{{< pagebreak >}}

![Likelihood of exceeding regulatory thresholds in each bay segment under different loading conditions and salinity changes.  Results for two time periods are shown where starting conditions were those from 1992 to 1994 (red) and from 2017 to 2021 (blue).  Line show the average and standard deviations of likelihoods of exceeding the thresholds for 10,000 simulations with each GAM by bay segment. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/simplo.png){#fig-simplo}

{{< pagebreak >}}

![Example simulations of chlorophyll-a for Old Tampa Bay for the period 2017 to 2021.  One hundred simulations are shown for the first year where salinity is set at observed and those for expected salinity fifty years from present.  Nitrogen load values are set to observed.  Mean chlorophyll across the simulations are shown by the black loess smooth, whereas actual mean annual chlorophyll is shown by the blue line.  The regulatory threshold is shown by the horizontal red line.](figs/simex.png){#fig-simex}

{{< pagebreak >}}

## Tables {.unnumbered}

```{r}
#| label: tbl-gamtab
#| tbl-cap: !expr 'paste0("Generalized Additive Model (GAM) summary statistics for chlorophyll-a predictions by bay segment in Tampa Bay.  Results for the overall model fit are shown with the Generalized Cross Validation (GCV) score and deviance explained (Dev. Expl.).  Summary statistics for the smoothers used in each model are also shown with the estimated degrees of freedom (edf), reference degrees of freedom (Ref.df), F-statitic (F), and p-value (p, bold < $\\alpha$ = 0.05). `s()` indicates a single spline smooth for the predictor and `ti()` indicatees a tensor product smoother interaction between two predictors.  dec_time: decimal time; doy: day of year; sal: salinity (ppth); tn_load: total nitrogen load (tons/month). OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
gamtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-gamcrtab
#| tbl-cap: !expr 'paste0("Correlations between monthly observed and predicted chlorophyll-a for different annual and seasonal periods for each bay segment. The Annual column shows the correlation for the entire year grouping (1985-1999, 2000-2014, or 2015-2024) and the remaining columns show the seasonal correlations within the year group. Font shading and size is proportional to the strength of the correlations. See @tbl-gamtab for individual model summary statistics. JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September; OND: October, November, December. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
gamcrtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-prdnrmtab
#| tbl-cap: !expr 'paste0("Mean absolute differences ($\\mu$g/L) for monthly predictions minus salinity-normalized results (Salinity) and salinity-normalized results minus salinity-normalized results at mean loadings (Load). The values in the salinity columns can be considered the relative effects of changes in salinity on chlorophyll-a, whereas those in the load columns are relative effects of loading on chlorophyll-a independent of salinity. The Annual columns show the differences for the entire year grouping (1985-1999, 2000-2014, or 2015-2024) and the remaining columns show the seasonal differences within the year group. Font shading and size is proportional to the magnitude of the differences. JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September; OND: October, November, December. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
prdnrmtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-liktab
#| tbl-cap: !expr 'paste0("Likelihoods of exceeding regulatory chlorophyll-a thresholds for each bay segment by simulated time period and loading scenarios under projected future salinity changes.  Results shown are likelihoods with standard deviations and the relative changes from present for years 25 and 50 in parentheses. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
liktab
```

{{< pagebreak >}}

## Data Availability {.unnumbered}

All data and analysis code for this manuscript is available on GitHub at <https://github.com/tbep-tech/chlhydro-manu>.

## Acknowledgments {.unnumbered}

This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP). We are thankful for constructive comments provided by xxx anonymous reviewers. We are indebted to the community of field and technical staff that have routinely collected and maintained the valuable long-term monitoring data used in this study.

## Statements and Declarations {.unnumbered}

### Author Contributions {.unnumbered}

Conceptualization: Marcus W Beck, Kerry Flaherty-Walia, Maya C Burke, Edward T Sherwood; Data Curation: Marcus W Beck; Methodology: Marcus W Beck, Kerry Flaherty-Walia, Maya C Burke, Edward T Sherwood; Formal analysis and investigation: Marcus W Beck; Writing - original draft preparation: all authors; Writing - review and editing: all authors; Funding acquisition: Edward T Sherwood.

### Ethics Declaration {.unnumbered}

The authors have no competing interests to declare that are relevant to the content of this article.

### Funding {.unnumbered}

This study was funded by EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St. Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.

## References {.unnumbered}