---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Managing the unmanageable: Understanding the competing effects of salinity and nutrient loading on estuarine water quality"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
crossref: 
  fig-title: Fig.
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      orcid: 0000-0002-4996-0059
      institute: [TBEP]
      correspondence: true
  - Kerry Flaherty-Walia:
      email: kfwalia@tbep.org
      orcid: 0000-0002-6032-0570
      institute: [TBEP]
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      orcid: 0000-0002-3847-6501      
      institute: [TBEP]
      correspondence: false 
  - Edward T Sherwood:
      email: esherwood@tbep.org
      orcid: 0000-0001-5330-302X
      institute: [TBEP]
      correspondence: false

institute:
  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA

filters: 
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  
execute: 
  echo: false
  warning: false
  ft.keepnext: false
  ft.align: left
---

```{r}
#| results: hide
library(knitr)
library(flextable)
library(here)
library(dplyr)
library(tidyr)

source(here('R/funcs.R'))

# inline data
load(file = here('data/mods.RData'))
load(file = here('data/lddatsrc.RData'))

# tables
load(file = here('tabs/gamtab.RData'))
load(file = here('tabs/gamcrtab.RData'))
load(file = here('tabs/prdnrmtab.RData'))
load(file = here('tabs/liktab.RData'))
```

## Inroduction 

[@testa2025]

Nitrogen loads in Tampa Bay have been reduced substantially since the 1970s due to concerted efforts to quantify and manage external nutrient inputs.  

WRTDS vs GAMs

The objectives of this study were to (1) develop Generalized Additive Models (GAMs) for chlorophyll-a trends for the four major bay segments of Tampa Bay as a function of time, salinity, and total nitrogen loading, (2) quantify the relative influence of salinity changes and nitrogen loading on chlorophyll-a trends, and (3) assess the potential likelihoods of exceeding regulatory criteria under future salinity changes and different loading scenarios. We hypothesize that water quality drivers have changed over time, warranting a shift in thinking about management strategies to account for the effects of factors that are difficult to control (salinity) under conventional management paradigms that focus on controlling external nutrient loading. This shift may be particularly important in upper bay segments where long-term nitrogen loads have changed, residence time is affected by freshwater inflows and barriers to circulation, and current loading thresholds may be insufficient to protect water quality under future expected changes in salinity. Overall, this work provides an in-depth assessment of chlorophyll trends for informing the management of water quality under past and contemporary conditions, with a generalized framework that could be applied in other locations with long-term data.

## Methods

Tampa Bay, located on the west coast of Florida, is a sub-tropical shallow estuary covering 400 m$^2$ (1036 km$^2$) of open water that includes native habitats of seagrasses, macroalgae, oysters, tidal flats, and hard bottom ([@fig-map]).  Average depth of the bay is 11 ft with a maximum depth of 43 feet in the dredged shipping channels that primarily traverse the longitudinal axis. The water column is well-mixed and typically does not exhibit strong stratification due primarily to the shallow depth. The bay is divided into four major segments defined by physical and natural boundaries, with Old Tampa Bay (OTB) in the northwest, Hillsborough Bay (HB) in the northeast, Middle Tampa Bay (MTB), and Lower Tampa Bay (LTB) at the southernmost extent that exchanges with the Gulf of Mexico ([@fig-map]). The watershed is 2200 mi$2$ (5872 km$^2$) and includes an extensive human presence, with 42% of the land as urban or suburban [@SWFWMD25]. Agricultural and mining activities cover about 25% of the watershed, whereas the remainder includes native habitats of hardwood or scrub uplands and freshwater wetlands.  

Residence time varies with inflow conditions, wind, and spatial scale.  In general, baywide estimates range from 75 (Eulerian) to 159 (Lagrangian) days [@burwell00;@morrison11], with substantial variation by location. Spatially, major inflows enter HB via the Hillsborough and Alafia Rivers, MTB via the Little Manatee river, and LTB via the Manatee River. Comprehensive estimates of residence time by bay segment do not exist, although time generally increases from the mouth of the bay towards the more oligohaline segments. LTB that exchanges with the Gulf has the shortest residence time of a few days, whereas OTB and HB have the longest.  However, HB flushes relatively rapidly compared to OTB due to major freshwater inflows and enhanced baroclinic circulation via the main dredge channel that extends from the mouth of the bay to upper HB [@morrison11].  More importantly, OTB has three major bridges with substantial causeways that are known impediments to circulation [@luther2022].  Residence times in OTB have been estimated at roughly three months [@tbep1115].  Throughout, we consider salinity as a rough proxy for residence time, while also acknowledging that it reflects hydrologic inputs, tidal exchange, baroclinic circulation, and other factors that influence hydrodynanmics (e.g., persistent eddies, wind).

Long-term external nutrient loading from multiple sources has decreased overall in Tampa Bay over the last fifty years [@janicki23]. Point sources primarily from wastewater treatment facilities with outfalls to the bay or receiving surface waters have been substantially reduced following upgrades to advanced tertiery treatment, resulting in an overall reduction of approximately two thirds from the 1970s to present for the entire bay. Conversely, nonpoint sources primarily from stormwater runoff remain the dominant source of nutrients to the bay, particularly in OTB and HB. Loading from nonpoint sources varies substantially between years with changes in rainfall and with more long-term climatic drivers (i.e., El Niño vs La Niña years).  However, a weak but long-term inrease in rainfall over the past fifty years has lead to gradual increases in stormwater loads and, consequently, a freshening of Tampa Bay, particularly in the upper bay segments [@beck2024]. Overall, HB receives the largest external nitrogen loads on an annual basis, followed by MTB, OTB, and LTB, respectively.

Reporting and management of external nutrient loads to Tampa Bay have been the responsibility of the Tampa Bay Nitrogen Management Consortium (TBNMC) since the early 2000s.  These efforts are guided by a Reasonable Assurance Plan whereby compliance assessments and water quality improvement efforts are reported to the Florida Department of Environmental Protection.  Regulatory criteria for chlorophyll-a apply to each bay segment, such that exceedances for more than two consecutive years require an assessment of a bay segment's assimilative capacity.  The assimilative capacity for each bay segment is currently defined by a maximum annual nitrogen loading beyond which water quality outcomes detrimental to seagraasses are expected. Similarly, chlorophyll thresholds for each bay segment were established to maintain a light environment supportive of seagrass growth.  Efforts of the TBNMC to manage external nutrient loads have contributed to the long-term recovery of Tampa Bay.  However, current ecology and physical conditions of the bay compared to past conditions from which regulatory thresholds were developed have been the focus of recent discussions by the TBNMC and have also motivated this current work.  Chlorophyll-a and total load thresholds for each bay segment are evaluated for objective three using the GAMs for scenario assessments under expected future changes and management actions.

### Data processing

Long-term monitoring data for Tampa Bay were obtained from the Environmental Protection Commission of Hillsborough County.  This monitoring program has been ongoing since the mid-1970s, with monthly sampling at fixed stations throughout the bay ([@fig-map]). Surface chlorophyll-a and salinity observations at each station were averaged by month for each bay segment, with salinity averaged across bottom, mid, and surface depth measurements at each station.  Salinity and chlorophyll gaps were filled using linear interpolation between observed values, where the maximum gap filled was three months.  Salinity and chlorophyll-a gaps accounted for less than 5% and 2%, respectively, of the observations for each bay segment. Monthly data from 1985 to present by bay segment were used for analysis to coincide with the period of record for the nutrient loading data (n = 480 for each bay segment). 

Monthly total nitrogen loading data to each bay segment from 1985 to present were estimated following methods in @tbep0494 [additional details in @tbep0623].  In brief, loading data were estimated as tons per month by external source, including domestic point-sources (wastewater treatment effluent, reclaimed water), industrial point sources, non-point sources, atmospheric deposition, groundwater/springs, and material losses from nutrients entering the bay through port/shipping activities.  Point source loads were estimated from reported flow and concentration data for every permitted entity in the Tampa Bay watershed. Conversely, non-point source loads were estimated from gauged and ungauged portions of the watershed by combining rainfall and flow estimates with event mean concentration data by land use and soil type.  Atmospheric deposition was estimated using rainfall and airshed chemical constituent data obtained from the Verna well-field long-term air quality monitoring site.  Groundwater loading was estimated from aquifer concentration data combined with potentiometric sub-surface flow data, whereas springs loadings were calculated directly from discharge and concentration data measured at each spring.  Finally, material losses were based on a loss factor per unit shipped at each port location.  For the purpose of this analysis, no distinction was made between load sources, such that the total load as the sum across sources was used for modelling.  However, we note that trends by sources have varied by location, e.g., non-point source loads have increased in OTB, whereas point sources have decreased [@janicki23].

### GAMs

```{r}
vifs <- mods |> 
  mutate(
    vif = purrr::map(data, function(x){
        mod <- lm(
          chla ~ doy + dec_time + sal + tn_load,
          data = x
        )
      car::vif(mod)
    }))
  
```
Four separate GAMs were created for each bay segment using the combined water quality and loading data for the monthly observations from 1985 to present.  The primary goal was to develop the best fitting model for observed chlorophyll-a by including the effects of salinity, nitrogen loading, and time.  Time was expressed as two separate variables for each time step, first as continuous decimal time (e.g., July 1st 2017 is 2017.5) and as day of year to capture a seasonal component.  Each predictor was evaluated as an additive smooth to quantify an individual assocation with chlorophyll-a, whereas interaction smoothers were also included for every pairwise combination of the predictors. No efforts were made for variable selection since the same model structure for describing drivers and scenario testing between bay segments was desired (objectives 2 and 3).  This also allowed direct comparability between models regarding explained variance, either as a whole or for individual predictors.

The model notation was:

$$
\begin{align}
\text{chla}_i &\sim \text{Gamma}(\mu_i, \phi) \\
\log(\mu_i) &= \beta_0 + f_1(\text{dec\_time}_i) + f_2(\text{doy}_i) + f_3(\text{sal}_i) + f_4(\text{load}_i) + \\
&\quad + f_{12}(\text{dec\_time}_i, \text{doy}_i) \\
&\quad + f_{13}(\text{dec\_time}_i, \text{sal}_i) \\
&\quad + f_{14}(\text{dec\_time}_i, \text{load}_i) \\
&\quad + f_{23}(\text{doy}_i, \text{sal}_i) \\
&\quad + f_{24}(\text{doy}_i, \text{load}_i) \\
&\quad + f_{34}(\text{sal}_i, \text{load}_i)
\end{align}
$$

where chlorophyll-a is modelled as a Gamma-distributed response variable with mean $\mu_i$ and dispersion parameter ϕ for observation $i$ (by month).  The decision to use a Gamma distribution as compared to a Gaussian with a log-link function was made given that the variance likely increases with the mean, which is a reasonable expectation for chlorophyll data. The log-transformed mean chlorophyll-a is modelled as an intercept ($\beta_0$) plus smooth functions of decimal time (dec_time), day of year (doy), salinity (sal), and loading (load), along with tensor product smooth interactions between each pair of the four predictors.  Specifically, $f_1​()$ is a thin plate regression spline with k = 40 basis functions, $f_2()$ is a cyclic cubic regression spline with k = 10 basis functions, $f_3()$ and $f_4()$ are thin plate regression splines with k = 10 basis functions, and $f_{12}()$, $f_{13}()$, $f_{14}()$, $f_{23}()$, $f_{24}()$, and $f_{34}()$ are tensor product smooths (excluding main effects) with k = (5, 5) basis functions.  Smoothing parameters were estimated using restricted maximum likelihood [@wood11] using the `gam()` function from the mgcv R package [@wood17].  Collinearity among model predictors was evaluated with variance inflation factors (VIFs), all of which were sufficiently low for inclusion [all VIF < 2, @fox2018]. 

The K values (number of basis functions) for each smoother were chosen to allow sufficient degress of freedom to describe the assocation of each predictor with chlorophyll.  Once models were fit, the `gam.check()` function was used to compare the estimated degrees of freedom for each smoother to the maximum degrees of freedom allowed by the K value.  None of the smoothers were within 80% of the maximum degrees of freedom allowed, suggesting sufficient flexibility to describe the assocation. ACF plots [@venables2002b] also showed minimal autocorrelation of model residuals at lag 1 (all < 0.20 $\rho$), indicating a time-correlated variance structure was not needed for the models.

In addition to predicted chlorophyll-a, the GAMs were used to create two additional types of predictions.  First, salinity-normalized predictions were made by predicting chlorophyll-a across a range of salinity values (minimum to maximum observed) at each time step, then taking the mean of those predictions as the salinity-normalized value for that time step.  This produces a time series that can be considered the chlorophyll trend independent of changes in salinity (i.e., just time and loading).  The approach is consistent with other studies examining the effects of salinity on chlorophyll-a concentrations [@beck17;@murphy19].

Salinity-normalized predictions at mean loading were also estimated.  These predictions can be considered the chlorophyll trend independent of changes in both salinity and loading.  This is accomplished by predicting chlorophyll-a at each time step using the mean loading with the salinity-normalized predictions for each bay segment across the entire time series.  Mean loading is used in this case as opposed to the more involved grid based approach for salinity normalization.  This was preferred for ease of interpretation and absence of a typical seasonal cycle that would be expected for a variable like salinity.

Finally, the models for each bay segment were evaluated for fit using conventional GAM summary statistics. These statistics included Generalized Cross-Validation scores and explained deviance, as well as estimated and reference degrees of freedom, F, and p-values for each smoother.  Simple correlations between observed and predicted chlorophyll were also evaluated for different year groupings (1985-1999, 2000-2014, 2015-2024) and seasonal periods (annual, JFM, AMJ, JAS, OND) to assess when model fit may differ relative to the entire time series.

### Partitioning Drivers of Chlorophyll Change

The second objective was to assess the relative effects of salinity and loading on chlorophyll trends from the GAMs for each bay segment. This was accomplished by comparing the predicted values from the GAMs with those from the salinity-normalized predictions and the salinity-normalized predictions at mean loading. Specifically, the difference between the predictions and salinity-normalized values as a simple subtraction was used an expression of salinity effects on chlorophyll.  Likewise, the difference between the salinity-normalized values and the salinity-normalized values at mean loading was used an expression of the loading effects on chlorophyll.  

The differences among the predictions by bay segment were assessed as simple visualizations of the annual results over time and using a state-space diagram plotting the differences for each time series over time on separate axes.  For the latter, the x-axis showed the salinity effects (positive being more oligohaline, negative being more euhaline) and the y-axis showed the loading effects (positive being high loading associated with higher chlorophyll, negative being lower loading associated with lower chlorophyll).  Individual years were plotted on the diagram to assess the relative effects of salinity changes and/or loading on the relative chlorophyll changes each year.  Depending on where a year is located on the plot, conclusions can be made on which factor (or both) may or may not have influenced chlorophyll each year.  Additionally, these differences were further evaluated for different year groupings (1985-1999, 2000-2014, 2015-2024) and seasonal periods (annual, JFM, AMJ, JAS, OND) to further evaluate periods of time when salinity and/or loading had differing effects on chlorophyll.

### Likelihood of Exceeding Regulatory Criteria

The third objective was to use the GAMs to assess the likelihood of exceeding regulatory criteria under projected (future) salinity changes and loading scenarios.  Generally, this was accomplished by repeated simulations from each GAM for the different scenarios.  The `simulate()` function from the gratia R package [@simpson25] was used, which predicts randomized GAM outputs within the prediction interval of the model.  For each scenario, 10,000 simulations were used to estimate chlorophyll, where the mean (and standard deviation) of the percent of years exceeding the thresholds were used to identify the likelihood of exceeding regulatory criteria over time.  Herein, "over time" is defined as a hypothetical period from present to fifty years in the future at one year time steps based on potential salinity changes and loading values for each bay segment. Chlorophyll thresholds for each bay segment are 9.3, 15.0, 8.5, and 5.1 $\mu$g/L and external loading thresholds are 486, 1451, 799, 349 tons per year for OTB, HB, MTB, and LTB, respectively. These thresholds were defined as being supportive of seagrass by promoting a sufficient light environment for growth [@tbep0696;@tbep0400;@tbep0310].

Salinity changes were estimated using the slopes from linear regressions from 1975 to present for the annual average salinities in each bay segment, i.e., to obtain estimated change in ppth per year.  Loading scenarios were assessed under four different conditions - current load, loading at the TMDL threshold, half the threshold, and twice the threshold.  The simulations and scenarios were applied to the 2015-2024 period of record for each bay segment. Changes in salinity were implemented for the chosen time period at one year time steps up to "50 years" using the expected change per year, where the observed salinity time series was multipled by the slope and the numbers of years in the "future" (one to 50).  Similarly, loading scenarios were evaluated for the four different conditions described above using the observed loading or scaled accordingly within each year, such that the sum of monthly loadings each year was equal to the annual sum for each relative scenario (i.e., half the TDML, equal to TMDL, or double the TMDL).  Practically, the results can be interpreted as the relative effects of potential salinity and loading changes if conditions within the chosen year period were to persist for fifty years into the future.

## Results

### Model Performance

```{r}
gamcrseas <- gamcrtab$body$dataset |>
  pivot_longer(names_to = 'seas', values_to = 'cr', JFM:OND) |>
  summarise(
    cr = round(mean(cr), 2),
    .by = seas
  )
gamcrseas <- split(gamcrseas, gamcrseas$seas)
```

GAMs fit for each bay segment had generally high explained deviance for observed chlorophyll, although differences were observed (@fig-obsprd, @tbl-gamtab).  All bay segments had overall explained deviance of at least 70% (OTB 74%, HB 70%, MTB 76%), whereas LTB had the lowest at 60%.  Model predictions over time captured the seasonality of chlorophyll and substantial variation in the seasonality was observed between years.  The long-term decreases in chlorophyll were notable in the upper bay segments, i.e., purple lines earlier in the time series generally higher chlorohpyll-a predictions than the green and yellow lines later in the time series for OTB and HB ([@fig-obsprd]b).  All bay segments had significant ($p$ < 0.05) individual smoothers for decimal time, day of year, and salinity, whereas only LTB had a significent smoother for nitrogen loading [@tbl-gamtab].  However, every bay segment had at least one significant interaction smoother for nitrogen loading suggesting an effect of loading on chlorophyll-a that was dependent on values of other predictors (loading and decimal time for OTB; loading and decimal time, day of year, and salinity for HB; and loading and day of year for MTB and LTB). 

Differences in correlations between observed and predicted chlorophyll were also shown by annual groupings and seasonal periods (@tbl-gamcrtab), suggesting fit varied by time in each bay segment.  Annual groupings by bay segment had correlations greater than 0.7, although LTB had slightly lower correlations than the other bay segments.  The highest annual grouping correlation was observed for 2015 to 2024 for OTB ($\rho$ = 0.90), whereas the lowest was observed for 2000 to 2014 for LTB ($\rho$ = 0.73).  By season, the October, November, December (OND) period generally had higher correlations across annual groupings and bay segments (mean $\rho$ = `r gamcrseas$OND$cr`), with the exception of OND for the 2000 - 2014 and 2015 - 2024 annual groupings for LTB ($\rho$ = 0.43, 0.41, respectively). Mean correlations were similar between bay segments and annual groupings for all other seasons (mean $\rho$ = `r gamcrseas$JFM$cr`, `r gamcrseas$AMJ$cr`, `r gamcrseas$OND$cr` for JFM, AMJ, and JAS, respectively).  The lowest seasonal correlation was observed for the January, February, March period for OTB in 2000-2014 ($\rho$ = 0.22).   The highest seasonal correlation was observed for the OND period for OTB in 2015-2024 ($\rho$ = 0.92).

### Drivers of Chlorophyll Change

Annually-averaged model predictions showing the salinity-normalized and salinity-normalized results at mean loading provided a general description of the differing effects of salinity and loading on chlorophyll over time [@fig-prdnrm]. For each bay segment, the plots were evaluated based on differences between the predictions (black points), salinity-normalized results (solid red lines), and salinity-normalized results at mean loading (dashed blue lines).  All bay segments except LTB had normalized results (salinity and salinity at mean loading) that differed largely from model predictions, suggesting that salinity and/or loading were key drivers of chlorophyll-a that varied over time.  For LTB, the normalized results generally followed the predictions, suggesting a relatively lower influence of salinity and loading on chlorophll-a in this bay segment. Further, OTB and HB had salinity-normalized results that differed from the results at mean loading, whereas the results for MTB and LTB had salinity-normalized results that were similar to those at mean loading.  This result suggested that loading had more unique effects on chlorophyll-a in OTB and HB, whereas loading was less important for MTB and LTB. The unique effects of loading in OTB and HB were most notable early and later in the time series, as indicated by larger deviations of the solid red lines from the dashed blue lines.  Comparing MTB ad LTB suggests that salinity had a larger influence on chlorophyll-a in the former, whereas loading was not a key driver in both.

The results in @fig-prdnrm were further assessed by the relative differences in the predictions, salinity-normalized results, and salinity-normalized results at mean loading by annual groupings and seasons [@tbl-prdnrmtab].  The results are expressed as mean absolute differences in chlorophyll for the three types of predictions in @fig-prdnrm to quantitavely describe the unique effects of salinity (predictions minus salinity-normalized results) and loading (salnity-normalizd results minus salinity-normalized results at mean loading) in the left and right side of @tbl-prdnrmtab, respectively.  As described above, salinity and loading had larger influences on chlorophyll-a for OTB and HB, salinity mostly influenced chlorophyll in MTB, and neither salinity nor loading had large influences in LTB.  Annually, salinity had the largest influence in OTB, HB, and MTB from 1985-1999, whereas loading was most important in OTB and HB from 2015-2024, although the loading effect was smaller compared to the salinity effect in OTB during 2015-2024.  By season, salinity effects were most prominent during the June, July, and August (JJA) periods in HB and MTB, whereas the OND period had the largest effect of salinity in OTB.  Salinity effects for JJA for HB and MTB and for OND for OTB were largest from 1985-1999. Loading effects by season were most prominent during the JJA period in OTB and HB, with the largest seasonal effects for JJA observed during 2015-2024.

The state-space representation provided a complementary interpretation of the effects of salinity and loading on chlorophyll-a over time [@fig-salvload].  Differences in scale for each plot axis by bay segment provided a relative comparison of the importance of salinity and loading, i.e., OTB and HB had the largest scales, MTB had a small scale for loading but not salinity, and LTB had small scales for both salinity and loading.  Overall by annual grouping (1985-1999, 2000-2014, and 2015-2024), OTB and HB suggested a shift towards more load-dominated influence on chlorophyll-a in 2015-2024 as shown by a shift in the mean effects higher on the y-axis ([@fig-salvload]b).  A shift towards the right on the x-axis for the 2015-2024 grouping suggested loads were also influenced by more oligohaline conditions, but to a lesser degree than the loading effects. Although general trends by annual groupings were shown by the mean effects in [@fig-salvload]b, substantial inter-annual variability was shown for some years ([@fig-salvload]a).  For example, exceptionally wet years that had a disproportionate effect on chlorophyll-a were seen as outliers in 1998 on the x-axis for OTB, HB, and MTB.  Similarly, years with higher than normal loading effects were seen as outliers in 2024 on the y-axis for OTB and HB.

### Likelihood of Exceeding Regulatory Criteria

Scenario assessments for recent years (2015-2024) suggested future reductions in salinity will contribute to increased likelihoods of exceeding the respective chlorophyll-a thresholds for each bay segment (@tbl-liktab).  However, these effects were trivial for all bay segments except OTB (describd below) because thresholds have not been exceeded in recent years.  That is, reductions in salinity in HB, MTB, and LTB suggested an increased likelihood although overall likelihoods were minimal (< 10% likelihood over 50 years).  One exception is HB when the loading was doubled, such that the likelihood of exceeding the regulatory threshold rose from ~1% at the current load to 24.9% at present and 28.4% in fifty years with reduced salinity.  These increased likelihoods are still marginal compared to OTB.  

The remainder of this section focuses on the likelihood that OTB will exceed regulatory criteria under projected salinity changes and different loading scenarios. Because actual regulatory exceedences were observed from 2015-2024 (five years in this period), the likelihood at present under observed salinity and loading was approximately 42% ([@tbl-liktab]).  Any changes in salinity and loading had non-trivial effects on the overall likelihood.  As an example, a subset of 100 simulations showed the effects of changing salinity for each loading scenario on predicted annual chlorophyll-a relative the regulatory criteria of 9.8 $\mu$g/L ([@fig-simplo]a). The left column in [@fig-simplo]a shows the results for present salinity and the right column shows the results for salinity fifty years from present (at a decrease of 0.04 ppth per year), where the rows (from top to bottom) show results for the actual load, half the TMDL loading, TMDL loading, and twice the TMDL loading.  The solid black lines show the average chlorophyll-a across the simulations, relative to the actual chlorophyll (solid blue lines) for each scenario.  The decrease in salinity resulted in higher predicted chlorophyll-a for all loading scenarios.  As expected, higher loading conditions also resulted in higher chlorophyll.  

Results across all years (1 year to 50) for the full 10,000 simulations show the overall likelihood of exceeding the regulatory threshold in OTB under the different scenarios ([@fig-simplo]b). If current loadings are maintained, the likelihood of exceeding the regulatory thresholds increases by 17% by year 50 to ~60%.  Relative increases in the likelihood are similar for each loading scenario, although the curves shift up or down proportionally depending on the loading.  Understandably, reducing the load by half shifted the likelihood curves down, whereas doubling the load shifted the curves higher.  Importantly, maintaining loading at the TMDL (slightly lower than the observed average loading from 2015-2024) showed that the likelihood of exceeding the regulatory criteria increases from 36% to 56% by year 50 with expected changes in salinity.

## Discussion

```{r}
# get slope of change per year by source, baysegment
lddatsrc_slopes <- lddatsrc |>
  dplyr::summarise(
    slope = coef(lm(tn_load ~ yr))[2],
    .by = c(bay_segment, source)
  )
strmslo <- lddatsrc_slopes |>
  dplyr::filter(source == 'NPS' & bay_segment == 'OTB') |>
  dplyr::pull(slope) |>
  as.numeric() |>
  round(1)

# get five year avg load by bay segment
lddat_avgload <- lddatsrc |>
  dplyr::filter(yr >= 2017 & yr <= 2021) |>
  dplyr::summarise(
    avgload = sum(tn_load) / 5,
    .by = bay_segment
  )
otb_avgload <- lddat_avgload |>
  dplyr::filter(bay_segment == 'OTB') |>
  dplyr::pull(avgload) |>
  as.numeric() |>
  round()
```

The underlying management paradigm for Tampa Bay emphasizes the control of external nutrients to create a light environment that is supportive of seagrass.  This paradigm has effectively supported the long-term recovery of seagrass in Tampa Bay, whereas shifting baselines, both present and in the future, may diminish the effectiveness of this approach [@sherwood2017;@beck2024].  The management paradigm for Tampa Bay was conceived over thirty years prior based on conditions of the Bay in the 1990s [@tbep0696;@tbep0400].  It is understood that conditions have changed and the primary objective of this work was to assess the time-varying effects of loading and salinity on water quality in Tampa Bay.  Our results demonstrated that chlorophyll-a conditions are affected by both, particularly in the upper bay, and that the dominant drivers have shifted over time.  This information has ecological importance by demonstrating how shifting baselines will continue to alter key drivers of water quality that can influence the ecosystem as a whole.  This information also has regulatory and managament implications. Mitigating external nutrient loads has been a fundamental approach for estuarine restoration, whereas salinity controls typically are not.  In anticipation of future salinity conditions, existing management paradigms may not be as effective as in the past to maintain water quality.  

The simulation results suggested an increased likelihood of exceeding regulatory thresholds with future reductions in salinity.  These analyses also underscored the continuing role that nutrient loads have for affecting water quality.  In OTB, where regulatory exceedances have occurred in several past years, changing salinity conditions may require further load reductions beyond the federally-established TMDL to maintain water quality in the future. Even if the nutrient TMDL is maintained (it is currently exceeded by `r otb_avgload - 486` tons for the last assessment period based on a five year average), the likelihood that the chlorophyll-a threshold will be exceeded increases by 9% in 25 years and by 20% in fifty years to an overall likelihood of 56%. These increases must also be considered relative to potential future loading trends if no actions are taken to mitigate sources. Although total loads across all sources in OTB have been relatively stable since 1985 (with subsantial inter-annual variability), contributions from non-point sources (stormwater primarily) have increased at a rate of `r strmslo` tons per year.  These load increases, if not addressed, will further increase the likelihood of exceeding regulatory criteria in the future.  The simulation models suggest that current total loading should be reduced by at least 50% to achieve a similar likelihood at present of exceeding the regulatory criteria in fifty years to account for the largely unmanageable effects of reduced salinity on chorophyll-a.

Model predictions compared to salinity-normalized and salinity-normalized results at mean loading suggested that chlorophyll in 1) OTB and HB was influenced by both salinity and loading, 2) MTB was influenced primarily by salinity, and 3) LTB was influenced by neither.  These relationships were also variable over time, particularly in OTB and HB, where pure loading effects appear to have recently increased in importance. This result has management implications for both bay segments, particularly for OTB where exceedances of the chlorophyll-a threshold have been observed in recent years.  Compliance assessment for the nutrient loadings in each bay segment was established as a loading ratio, where loads are scaled based on hydrologic inputs [@tbep0310]. This allows flexibility during wet years when loading is higher than normal so that stormwater managers are not unnecessarily penalized. For all bay segments except OTB, there has been agreement between meeting the chlorophyll threshold and the nutrient load ratios. Recent exceedances of the chlorophyll threshold in OTB and non-attainment of the loading ratio presents a mismatch in compliance assessment.  Because the nutrient loading ratios are based on hydrologic conditions from a 1992-1994 reference period, changing hydrology and rainfall patterns suggest this approach no longer describes a meaningful ecological relationship [@beck2024;@wang2026].  Attainment based on absolute loadings would be a more meaningful assessment, at least for OTB.

The model results were interpreted as the unique effects of both nutrient loading and salinity, one or the other, or neither on chlorophyll-a.  Although loading was suggested as having less of an influence on chlorophyll-a in MTB and LTB, the results do not imply loading is unimportant in these bay segments.  Rather, the results must be interpreted relatively between bay segments such that a general conclusion can be made that loading is more important for driving chlorophyll-a in the upper segments, not that is it wholly unimportant for the lower bay segments.  The GAM structure was selected specifically to isolate the effects of key drivers and is a relatively simplistic interpretation compared to a more mechanistic descriptions of factors driving chlorophyll-a. Controlling external nitrogen loads to each bay segment will continue to be a key focus for managements efforts to maintain or improve water quality in Tampa Bay. 

Recognizing what salinity represents as a proxy of causal mechanisms of chlorophyll dynamics in each bay segment is critical for the interpretation of the results.  Salinity is not independent from loading such that nutrient inputs to the bay are largely dependent on annual rainfall. Extreme wet or dry years have coincided with high and low nutrient loading years, which are prominent as the annual outliers in @fig-salvload.  From a statistical perspective, loading and salinity are correlated but not so excessive as to prevent both from being used as unique predictors in the models (VIFs were sufficiently low). Given the varying importance of either in the models for each bay segment, salinity can be interpreted as having a unique effect on chlorophyll independent of loading. The most likely interpretation is that salinity is a proxy for flushing, residence time, and tidal exchange [e.g., @hagy2000;@meyers2015].  Lower salinity not only suggests greater freshwater inputs, but is also likely associated with higher residence time with reduced flushing and tidal exchange.  Such periods of time could be visually interpreted from @fig-prdnrm when the solid red lines showing the salinity-normalized values are higher than the salinity-normalized values at mean loading.  The opposite could be true during dry periods or when flushing and tidal exchange is high (i.e., solid lines below the dashed blue lines).

Our approach offers a relatively novel application of GAMs to describe long-term trends in chlorophyll and how drivers have changed.  We build off similar past models used for these purposes [@richards2013;@murphy19;@beck2022b], although most of the existing applications have focused on trend description (e.g., how has chlorophyll-a changed?) and not describing changing causal mechanisms (e.g., why has chlorophyll-a changed?). GAMs have had far less applications in the latter context for long-term monitoring data [but see, @murphy2021;@schramm2023].  The use of GAMs to simulate chlorophyll trends under different scenarios is also novel and highlights a strength of these models for evaluating future conditions or management interventions.  Other similar but functionally different methods (e.g., WRTDS models) cannot be readily used for these purposes.  Mechanistic models have also traditionally filled these roles, particularly for developing TMDLS in impaired waters [@borsuk2002;@wool2003;@borah2006] While mechanistic models will continue to serve an important role in this capacity, GAMs are more easily adapted to support management, either through modification of the core model structure or for scenario assessment.  

Moreover, GAMs offer unique opportunities for data visualization to support resource management.  For example, @fig-gridplo provides a clear depiction of the model results that helps develop an understanding of how chlorophyll has changed on annual and seasonal periods under different salinity regimes.  For added context, regulatory thresholds can also be shown on these plots to identify periods of time and conditions when criteria are likely to be exceeded.  For example, chlorophyll exceedances in OTB are most common during the summer under more oligohaline conditions.  Seasonally, exceedances have also been more infrequent in HB in the late summer/early fall, particularly in recent years in October when salinity has been lower.  Although regulatory compliance is based on annual chlorophyll averages, these visualizations can aid resource managers in identifying periods of time and conditions when water quality may be most vulnerable to additional loading. Most resource managers lack sufficient quantative tools and visualization methods to aid in understanding how external factors could affect the likelihood of exceeding regulatory thresholds.  These tools can help support proactive management interventions when future conditions are expected to increase the chance of being out of compliance.  

Finally, the scenario assessments using simulated results from the GAMs also assumed that current chloropyll standards are and will continue to support a light environment for seagrasses.  These chloropyll thresholds, much like the loading thresholds, were developed with data from the 1990s and current conditions suggest analysis is needed to determine if they are still appropriate. Optical models, in situ light measurements, and current seagrass coverage maps cross-referenced with bathymetry need to be leveraged to determine if these thresholds are still sufficient to maintain light environments at the deep edge of seagrass for each bay segment [@tbep0622]. Evaluating other water quality factors that attenuate light may also be important, such as color and turbidity [@chen2007;@le2013].  Although there are no standards to support the light environment other than chlorophyll, evaluating the roles of alternative drivers of light attenuation can help development expectations of times when seagrasses may be vulnerable.  For example, high rainfall events, such as during tropical storms, can elevate dissolved organic carbon that alters water color [@avery2004;@chen2007b].  The effects of these events on light attenuation and seagrass growth are not well understood in Tampa Bay. 

### Conclusions

This study evaluated the competing effects of nutrient loading and salinity on chlorophyll dynamics in the four major bay segments of Tampa Bay. Dominant drivers have changed over time, particularly for the upper bay segments (OTB, HB), which has relevance when considering that current management paradigms for the bay were conceived over thirty years ago when conditions were much different than those at present. We further demonstrated how future salinity changes, which are difficult to manage, will lead to higher likelihoods of exceeding regulatory criteria in the future, such that current loading thresholds may be insufficient to protect water quality.  Specifically, nutrient loads in Old Tampa Bay may need to be reduced by half of the current TMDL to maintain a sufficiently low likelihood of exceeding regulatory standards.  These models were developed only in the context of nutrient loading and salinity, and it is recognized that additional system-wide changes may further stress conditions in the Bay.  In particular, increased temperatures from climate change will not only physically stress the Bay's resources [e.g., seagrass growth and mortality, @beck2024], but will also alter physiological response of phytoplankton and nutrient cycling rates [@lassen2010;@statham2012a].  Future work should evaluate the role of additional stressors, like temperature, to better guide how Bay management should proceed to protect critical resources. Our approach can also be leveraged in other estuarine systems where the effectiveness of conventional managment paradigms should be assessed relative to changing baselines.

{{< pagebreak >}}

## Figures {.unnumbered}

![Tampa Bay on the west coast of Florida, USA.  Bay segments are denoted with the labels and their boundaries.  Points show long-term monitoring stations from the Environmental Protection Commission of Hillsborough County where chlorophyll-a and salinity data were used for modeling. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/map.png){#fig-map}

{{< pagebreak >}}

![Observed and predicted chlorophll-a ($\mu$g/L) for the four major bay segments of Tampa Bay.  Observed and predicted are shown continuously over time (a) and by month (b). Color shading indicates year. Rows are arranged by bay segment with the top labels in the left column showing the bay segment and the right column showing deviance explained.](figs/obsprd.png){#fig-obsprd}

{{< pagebreak >}}

![Annual predictions, normalized estimates, and normalized estimates at mean loading for chlorophyll-a by major bay segment from the separate GAMs. Normalized values (solid lines) are those that remove the salinity influence using the mean result of predictions across a range of salinity values at each time step, whereas the normalized, mean load values (dashed lines) are salinity-normalized values at a constant mean load for each bay segment. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/prdnrm.png){#fig-prdnrm}

{{< pagebreak >}}

![State space representations of modelled annual chlorophyll response per unit change in salinity (x-axis) or nitrogen load (y-axis).  Points represent values for a given year in (a), joined sequentially through time by the grey lines. Means and 95% confidence for the year groups are shown in (b). OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/salvload.png){#fig-salvload}

{{< pagebreak >}}

![Simulated chlorophyll for a subset of 100 simulations (a) and likelihood of exceeding regulatory thresholds with 10,000 simulations (b) in OTB under different loading conditions and salinity changes.  The left column in (a) shows simulated chl-a for the first year where salinity is set at observed and the right column shows results for expected salinity fifty years from present for the time period from 2015-2024.  Nitrogen load values were set to the actual observed load (top row), half the TMDL (second row), the TMDL (third row), and double the TMDL (bottom row). Lines in (b) show the average and standard deviations of likelihoods of exceeding the thresholds over time for 10,000 simulations using the OTB model.](figs/simplo.png){#fig-simplo}

{{< pagebreak >}}

![Predicted chlorophyll-a ($\mu$g/L) during the rainy season (June to November) from 2005 to present as related to salinity (ppth) by major bay segment.  Facets are arranged to show how the monthly predictions vary across years and different salnity ranges.  The points show the model predicted results for the observed salinity.  Salinity ranges in each plot are fixed to the seasonal minimum and maximum observed across years for the month. Grey lines show the chlorophyll contour value for the regulatory threshold for each bay segment. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/gridplo.png){#fig-gridplo}

{{< pagebreak >}}

## Tables {.unnumbered}

```{r}
#| label: tbl-gamtab
#| tbl-cap: !expr 'paste0("Generalized Additive Model (GAM) summary statistics for chlorophyll-a predictions by bay segment in Tampa Bay (n = 480 for each).  Results for the overall model fit are shown with the Generalized Cross Validation (GCV) score and deviance explained (Dev. Expl.).  Summary statistics for the smoothers used in each model are also shown with the estimated degrees of freedom (edf), reference degrees of freedom (Ref.df), F-statistic (F), and p-value (p, bold < $\\alpha$ = 0.05). `s()` indicates a single spline smooth for the predictor and `ti()` indicates a tensor product smoother interaction between two predictors.  dec_time: decimal time; doy: day of year; sal: salinity (ppth); tn_load: total nitrogen load (tons/month). OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
gamtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-gamcrtab
#| tbl-cap: !expr 'paste0("Correlations between monthly observed and predicted chlorophyll-a for different annual and seasonal periods for each bay segment. The Annual column shows the correlation for the entire year grouping (1985-1999, 2000-2014, or 2015-2024) and the remaining columns show the seasonal correlations within the year group. Font shading and size is proportional to the strength of the correlations. See @tbl-gamtab for individual model summary statistics. JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September; OND: October, November, December. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
gamcrtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-prdnrmtab
#| tbl-cap: !expr 'paste0("Mean absolute differences ($\\mu$g/L) for monthly predictions minus salinity-normalized results (Salinity) and salinity-normalized results minus salinity-normalized results at mean loadings (Load). The values in the salinity columns can be considered the relative effects of changes in salinity on chlorophyll-a, whereas those in the load columns are relative effects of loading on chlorophyll-a independent of salinity. The Annual columns show the differences for the entire year grouping (1985-1999, 2000-2014, or 2015-2024) and the remaining columns show the seasonal differences within the year group. Font shading and size is proportional to the magnitude of the differences. JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September; OND: October, November, December. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
prdnrmtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-liktab
#| tbl-cap: !expr 'paste0("Likelihoods of exceeding regulatory chlorophyll-a thresholds for each bay segment for different loading scenarios under projected future salinity changes.  Results shown are likelihoods of exceeding the threshold from 10,000 simulations from each model for the 2015-2024 time period.  Standard deviations and the relative changes from present for years 25 and 50 are shown in parentheses. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
liktab
```

{{< pagebreak >}}

## Data Availability {.unnumbered}

All data and analysis code for this manuscript is available on GitHub at <https://github.com/tbep-tech/chlhydro-manu>.

## Acknowledgments {.unnumbered}

This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP). We are thankful for constructive comments provided by xxx anonymous reviewers. We are indebted to the community of field and technical staff that have routinely collected and maintained the valuable long-term monitoring data used in this study.

## Statements and Declarations {.unnumbered}

### Author Contributions {.unnumbered}

Conceptualization: Marcus W Beck, Kerry Flaherty-Walia, Maya C Burke, Edward T Sherwood; Data Curation: Marcus W Beck; Methodology: Marcus W Beck, Kerry Flaherty-Walia, Maya C Burke, Edward T Sherwood; Formal analysis and investigation: Marcus W Beck; Writing - original draft preparation: all authors; Writing - review and editing: all authors; Funding acquisition: Edward T Sherwood.

### Ethics Declaration {.unnumbered}

The authors have no competing interests to declare that are relevant to the content of this article.

### Funding {.unnumbered}

This study was funded by EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St. Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.

## References {.unnumbered}