---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "An awesome title for a great paper"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
crossref: 
  fig-title: Fig.
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      orcid: 0000-0002-4996-0059
      institute: [TBEP]
      correspondence: true
  - Kerry Flaherty-Walia:
      email: kfwalia@tbep.org
      orcid: 0000-0002-6032-0570
      institute: [TBEP]
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      orcid: 0000-0002-3847-6501      
      institute: [TBEP]
      correspondence: false 
  - Edward T Sherwood:
      email: esherwood@tbep.org
      orcid: 0000-0001-5330-302X
      institute: [TBEP]
      correspondence: false

institute:
  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA

filters: 
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  
execute: 
  echo: false
  warning: false
  ft.keepnext: false
  ft.align: left
---

```{r}
#| results: hide
library(knitr)
library(flextable)
library(here)

source(here('R/funcs.R'))

load(file = here('tabs/gamtab.RData'))
load(file = here('tabs/gamcrtab.RData'))
load(file = here('tabs/prdnrmtab.RData'))
load(file = here('tabs/hydnrmtab.RData'))
```

## Fig

1. observed chl and model fit
1. model fit v normalized
1. grid plots
1. normalized - fit v hydrology

## Table

1. Model summary statistics (include by month?)

## Inroduction 

[@testa2025]

## Methods

### Data processing

Salinity gaps were filled using linear interpolation between observed values, where the maximum gap fill was three time steps (months).  Gaps accounted for less than 5% of the observatiosn for each bay segment.  

### GAMs

$$
\begin{align}
\text{chla}_i &\sim \text{Gamma}(\mu_i, \phi) \\
\log(\mu_i) &= \beta_0 + f_1(\text{dec\_time}_i) + f_2(\text{doy}_i) + f_3(\text{sal}_i) \\
&\quad + f_{12}(\text{dec\_time}_i, \text{doy}_i) \\
&\quad + f_{13}(\text{dec\_time}_i, \text{sal}_i) \\
&\quad + f_{23}(\text{sal}_i, \text{doy}_i)
\end{align}
$$

where chlorophyll-a is modelled as a Gamma-distributed response variable with mean $\mu_i$ and dispersion parameter ϕ for observation i.  The log-transformed mean chlorophyll-a is modelled as an intercept ($\beta_0$) plus smooth functions of decimal time (dec_time), day of year (doy), and salinity (sal), along with tensor product smooth interactions between each pair of the three predictors.  Specifically, $f_1​()$ is a thin plate regression spline with k = 40 basis functions, $f_2()$ is a cyclic cubic regression spline with k = 10 basis functions, $f_3()$ is a thin plate regression spline with k = 10 basis functions, and $f_{12}()$, $f_{13}()$, and $f_{23}()$ are tensor product smooths (excluding main effects) with k = (5, 5) basis functions.  Smoothing parameters were estimated using restricted maximum likelihood (REML).

ACF plots showed minimal autocorrelation of model residuals at lag 1 (all < 0.20 rho).

K values chosen to allow sufficient degress of freedom for each smoother.  Once models were fit, the `gam.check()` function was used to compare the estimated degrees of freedom for each smoother to the maximum degrees of freedom allowed by the k value.  None of the smoothers were within 80% of the maximum degrees of freedom allowed. 

Salinity-normalized predictions were made by predicting chlorophyll-a across a range of salinity values (minimum to maximum observed) at each time step, then taking the mean of those predictions as the salinity-normalized value for that time step.  

$$
\begin{align}
\text{Norm}(t) &= \frac{1}{|\mathcal{S}|} \sum_{s \in \mathcal{S}} \hat{f}\left(t, s, \bar{\text{sal}}_{m(t),y(t)}\right)
\end{align}
$$

### Scenario assessment

Make a note about forecasting with GAMs, they are tricky because the polynomial surfaces behave poorly outside of the range of the observed data.  

## Results

@fig-obsprd, @fig-prdnrm, @fig-gridplo, @fig-salvload, @fig-simplo, @fig-simex

@tbl-gamtab, @tbl-gamcrtab, @tbl-prdnrmtab, @tbl-hydnrmtab

## Discussion

Why do chlorophyll simulations using actual loads at time zero show lower likelihoods of exceeding the threshold than the actual percent of years in the time period that exceeded?  For example, in OTB from 2017-2021, 4 of the 5 years exceeded the threshold, producing an expectation that the baseline likelihood would be 80% for the actual load at time zero (observed salinity).  First, the actual values for chlorophyll were very close to the threshold.  Given the inherent uncertainty in measured data (observational and process uncertainy) and the uncertainty of the models (model uncertainty, affected by both observation and process uncertainty), a lower than expected likelihood of exceeding the threshold for that time period was provided by the simulations.  Further, annual exceedances are largely driven by high values during the summer months which produce a higher mean value in a variable that is inherently log-distributed.  Because GAMs, like most empirical models, describe the mean response, these high values are poorly described and are not common outputs in the simulated values.  So, the likelihoods of exceeding the threshold can be considered underestimates, whereas the more important result is the relative increase in the likelihood over time given projected salinity changes in the future.  In other words, all scenarios suggested greater exceedances of regulatory thresholds from present to fifty years.

{{< pagebreak >}}

## Figures {.unnumbered}

![Observed and predicted chlorophll-a ($\mu$g/L) for the four major bay segments of Tampa Bay.  Observed and predicted are shown continuously over time (a) and by month (b). Color shading indicates year. Rows are arranged by bay segment with the top labels in the left column showing the bay segment and the right column showing deviance explained.](figs/obsprd.png){#fig-obsprd}

{{< pagebreak >}}

![Annual predictions, normalized estimates, and normalized estimates at mean loading for chlorophyll-a by major bay segment from the separate GAMs. Normalized values (solid lines) are those that remove the salinity influence using the mean result of predictions across a range of salinity values at each time step, whereas the normalized, mean load values (dashed lines) are salinity-normalized values at a constant mean load for each bay segment. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/prdnrm.png){#fig-prdnrm}

{{< pagebreak >}}

![State space representations of modelled annual chlorophyll response per unit change in salinity (x-axis) or nitrogen load (y-axis).  Points represent values for a given year in (a), joined sequentially through time by the grey lines. Means and 95% confidence for the year groups are shown in (b). OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/salvload.png){#fig-salvload}

{{< pagebreak >}}

![Predicted chlorophyll-a ($\mu$g/L) during the rainy season (June to November) from 2005 to present as related to salinity (ppth) by major bay segment.  Facets are arranged to show how the monthly predictions vary across years and different salnity ranges.  The points show the model predicted results for the observed salinity.  Salinity ranges in each plot are fixed the minimum and maximum observed across years for the month. Grey lines show the chlorophyll contour value for the regulatory threshold for each bay segment. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/gridplo.png){#fig-gridplo}

{{< pagebreak >}}

![Likelihood of exceeding regulatory thresholds in each bay segment under different loading conditions and salinity changes.  Results for two time periods are shown where starting conditions were those from 1992 to 1994 (red) and from 2017 to 2021 (blue).  Line show the average and standard deviations of likelihoods of exceeding the thresholds for 10,000 simulations with each GAM by bay segment. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.](figs/simplo.png){#fig-simplo}

{{< pagebreak >}}

![Example simulations of chlorophyll-a for Old Tampa Bay for the period 2017 to 2021.  One hundred simulations are shown for the first year where salinity is set at observed and those for expected salinity fifty years from present.  Nitrogen load values are set to observed.  Mean chlorophyll across the simulations are shown by the black loess smooth, whereas actual mean annual chlorophyll is shown by the blue line.  The regulatory threshold is shown by the horizontal red line.](figs/simex.png){#fig-simex}

{{< pagebreak >}}

## Tables {.unnumbered}

```{r}
#| label: tbl-gamtab
#| tbl-cap: !expr 'paste0("Generalized Additive Model (GAM) summary statistics for chlorophyll-a predictions by bay segment in Tampa Bay.  Results for the overall model fit are shown with the Generalized Cross Validation (GCV) score and deviance explained (Dev. Expl.).  Summary statistics for the smoothers used in each model are also shown with the estimated degrees of freedom (edf), reference degrees of freedom (Ref.df), F-statitic (F), and p-value (p, bold < $\\alpha$ = 0.05). `s()` indicates a single spline smooth for the predictor and `ti()` indicatees a tensor product smoother interaction between two predictors.  dec_time: decimal time; doy: day of year; sal: salinity (ppth); tn_load: total nitrogen load (tons/month). OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
gamtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-gamcrtab
#| tbl-cap: !expr 'paste0("Correlations between monthly observed and predicted chlorophyll-a for different annual and seasonal periods for each bay segment. The Annual column shows the correlation for the entire year grouping (1985-1999, 2000-2014, or 2015-2024) and the remaining columns show the seasonal correlations within the year group. Font shading and size is proportional to the strength of the correlations. See @tbl-gamtab for individual model summary statistics. JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September; OND: October, November, December. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
gamcrtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-prdnrmtab
#| tbl-cap: !expr 'paste0("Mean absolute differences ($\\mu$g/L) for monthly predictions minus salinity-normalized results (Salinity) and salinity-normalized results minus salinity-normalized results at mean loadings (Load). The values in the salinity columns can be considered the relative effects of changes in salinity on chlorophyll-a, whereas those in the load columns are relative effects of loading on chlorophyll-a independent of salinity. The Annual columns show the differences for the entire year grouping (1985-1999, 2000-2014, or 2015-2024) and the remaining columns show the seasonal differences within the year group. Font shading and size is proportional to the magnitude of the differences. JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September; OND: October, November, December. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay.")'
prdnrmtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-hydnrmtab
#| tbl-cap: !expr 'paste0("Linear model fits between the predicted minus salinity-normalized chlorophyll and the hydrologic load anomalies by quarter, bay segment, and annual periods. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay. JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September; OND: October, November, December. p < 0.05 *; p < 0.005 **.")'
hydnrmtab
```

{{< pagebreak >}}

## Data Availability {.unnumbered}

All data and analysis code for this manuscript is available on GitHub at <https://github.com/tbep-tech/chlhydro-manu>.

## Acknowledgments {.unnumbered}

This work benefited from constructive discussions with the Technical Advisory Committee of the Tampa Bay Estuary Program (TBEP). We are thankful for constructive comments provided by xxx anonymous reviewers. We are indebted to the community of field and technical staff that have routinely collected and maintained the valuable long-term monitoring data used in this study.

## Statements and Declarations {.unnumbered}

### Author Contributions {.unnumbered}

Conceptualization: Marcus W Beck, Kerry Flaherty-Walia, Maya C Burke, Edward T Sherwood; Data Curation: Marcus W Beck; Methodology: Marcus W Beck, Kerry Flaherty-Walia, Maya C Burke, Edward T Sherwood; Formal analysis and investigation: Marcus W Beck; Writing - original draft preparation: all authors; Writing - review and editing: all authors; Funding acquisition: Edward T Sherwood.

### Ethics Declaration {.unnumbered}

The authors have no competing interests to declare that are relevant to the content of this article.

### Funding {.unnumbered}

This study was funded by EPA Section 320 Grant Funds, and the TBEP's local government partners (Hillsborough, Manatee, Pasco, and Pinellas Counties; the Cities of Clearwater, St. Petersburg, and Tampa; Tampa Bay Water; and the Southwest Florida Water Management District) through contributions to the operating budget.

## References {.unnumbered}