---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "*Supplement* Managing the unmanageable: Understanding the competing effects of salinity and nutrient loading on estuarine water quality"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
crossref: 
  fig-title: Fig.
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      orcid: 0000-0002-4996-0059
      institute: [TBEP]
      correspondence: true
  - Edward T Sherwood:
      email: esherwood@tbep.org
      orcid: 0000-0001-5330-302X
      institute: [TBEP]
      correspondence: false
  - Rebecca R Murphy:
      email: rmurphy@chesapeakebay.net
      orcid: 0000-0003-3391-1823
      institute: [CBP]
      correspondence: false
  - Kerry Flaherty-Walia:
      email: kfwalia@tbep.org
      orcid: 0000-0002-6032-0570
      institute: [TBEP]
      correspondence: false
  - Qian Zhang:
      email: qzhang@chesapeakebay.net
      orcid: 0000-0003-0500-5655
      institute: [CBP]
      correspondence: false
  - Blake A Simmons: 
      email: bsimmons@tbep.org
      orcid: 0000-0002-1918-3463
      institute: [TBEP]
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      orcid: 0000-0002-3847-6501      
      institute: [TBEP]
      correspondence: false 

institute:
  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA
  - CBP: University of Maryland Center for Environmental Science/U.S. Environmental Protection Agency Chesapeake Bay Program, 1750 Forest Drive, Annapolis, Maryland 21401 USA

filters: 
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  
execute: 
  echo: false
  warning: false
  ft.keepnext: false
  ft.align: left
---

```{r}
#| results: hide
library(knitr)
library(here)
library(flextable)

load(file = here('tabs/suppaictab.RData'))
load(file = here('tabs/suppgamtab.RData'))
load(file = here('tabs/suppconcurvotbtab.RData'))
load(file = here('tabs/suppconcurvhbtab.RData'))
load(file = here('tabs/suppconcurvmtbtab.RData'))
load(file = here('tabs/suppconcurvltbtab.RData'))
load(file = here('tabs/suppldtrendtab.RData'))
```

## Figures {.unnumbered}

![Examples demonstrating the salinity normalization method (a), the salinity normalization method at mean loading (b), and the final results (c).  Results are obtained by predicting chlorophyll-a concentrations from a fitted GAM across a range of salinities and estimating the average (a) or doing the same but using mean loading (b).  The bottom plot (c) shows the normalized curves where the red line is the mean curve from (a), the blue dashed line is the mean curve from (b), and the points are the actual model predictions.](figs/suppnorm.png){#fig-suppnorm}

{{< pagebreak >}}

![Effects plots for individual smoothers from Generalized Additive Models applied to each bay segment.  Each plot shows the modeled effects on chlorophyll-a of the smoothers identified on the x-axes.  The small vertical bars on the x-axis of each plot show the relative distribution of values for each parameter.  Confidence intervals (95%) are also shown for each parameter. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay](figs/suppgameff1.png){#fig-suppgameff1}

{{< pagebreak >}}

![Effects plots for bivariate smoothers from Generalized Additive Models applied to each bay segment.  Each plot shows the modeled effects on chlorophyll-a of the smoothers identified on the x- and y-axes.  Observed values are shown as points to emphasize the typical ranges for each parameter combination. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay](figs/suppgameff2.png){#fig-suppgameff2}

{{< pagebreak >}}

## Tables {.unnumbered}

```{r}
#| label: tbl-suppaictab
#| tbl-cap: "Comparison of Generalized Additive Models for chlorophyll-a for each bay segment using a Gamma and Gaussian (log-link) error distributions.  Model fit was assessed using Akaike information criterion [AIC, @akaike1974], with the lowest AIC value by segment shown in bold. df: degrees of freedom; OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay."
suppaictab
```

```{r}
#| label: tbl-suppgamtab
#| tbl-cap: "Generalized Additive Model (GAM) summary statistics for chlorophyll-a predictions by bay segment in Tampa Bay (n = 480 for each).  Results are shown for models with different total nitrogen loading lags, whether salinity or loading were corrected for seasonality, and constraints on the number of smoothers (K) for time, loading, and salinity (flexible or fixed). Loads were evaluated as cumulative sums for the current month up to a three month lag. Results for the overall model fit are shown with the Generalized Cross-Validation (GCV) scores in the top rows and deviance explained (Dev. Expl.) in the bottom rows. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay."
suppgamtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-suppconcurvotbtab
#| tbl-cap: "Concurvity estimates for the __Old Tampa Bay__ Generalized Additive Models. Values range from 0 to 1 and assess the degree of multicollinearity among predictors. Results are shown for models with different total nitrogen loading lags, whether salinity or loading were corrected for seasonality, and constraints on the number of smoothers (K) for time, loading, and salinity (flexible or fixed). Loads were evaluated as cumulative sums for the current month up to a three month lag. Font size and shading are proportional among cells to emphasize differences."
suppconcurvotbtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-suppconcurvhbtab
#| tbl-cap: "Concurvity estimates for the __Hillsborough Bay__ Generalized Additive Models. Values range from 0 to 1 and assess the degree of multicollinearity among predictors. Results are shown for models with different total nitrogen loading lags, whether salinity or loading were corrected for seasonality, and constraints on the number of smoothers (K) for time, loading, and salinity (flexible or fixed). Loads were evaluated as cumulative sums for the current month up to a three month lag. Font size and shading are proportional among cells to emphasize differences."
suppconcurvhbtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-suppconcurvmtbtab
#| tbl-cap: "Concurvity estimates for the __Middle Tampa Bay__ Generalized Additive Models. Values range from 0 to 1 and assess the degree of multicollinearity among predictors. Results are shown for models with different total nitrogen loading lags, whether salinity or loading were corrected for seasonality, and constraints on the number of smoothers (K) for time, loading, and salinity (flexible or fixed). Loads were evaluated as cumulative sums for the current month up to a three month lag. Font size and shading are proportional among cells to emphasize differences."
suppconcurvmtbtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-suppconcurvltbtab
#| tbl-cap: "Concurvity estimates for the __Lower Tampa Bay__ Generalized Additive Models. Values range from 0 to 1 and assess the degree of multicollinearity among predictors. Results are shown for models with different total nitrogen loading lags, whether salinity or loading were corrected for seasonality, and constraints on the number of smoothers (K) for time, loading, and salinity (flexible or fixed). Loads were evaluated as cumulative sums for the current month up to a three month lag. Font size and shading are proportional among cells to emphasize differences."
suppconcurvltbtab
```

{{< pagebreak >}}

```{r}
#| label: tbl-suppldtrendtab
#| tbl-cap: "Contemporary loading trends from 2012 to 2024 by bay segment.  Results are for regression models of annual loading with year. OTB: Old Tampa Bay; HB: Hillsborough Bay; MTB: Middle Tampa Bay; LTB: Lower Tampa Bay."
suppldtrendtab
```

{{< pagebreak >}}

## References {.unnumbered}